---
title: "STA465 hw3"
author: "Jiawei Gong"
date: "11/5/2024"
output:
  html_document: default
  pdf_document: default
editor_options: 
  markdown: 
    wrap: 72
---

#  {.tabset}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup

```{r}
library(tidyverse)
library(leaflet) 
library(sf)
library(units)
library(spdep)
library(spData) # for SIDS spatial data
library(spatialreg) # for SAR and CAR
library(spgwr) # for GWR
library(spaMM) 
```

```{r}
data.dir = gsub('/code', '/data', getwd())
neighbourhood_stats <- read.csv(paste0(data.dir, "/neighbourhood-stats-1.csv"))

neighborhoods= st_read(paste0(data.dir, '/Toronto_Neighbourhoods/Toronto_Neighbourhoods.shp'))
neighborhoods_wgs84 <- st_transform(neighborhoods, crs = 4326)
```

```{r}
#spatial join of crimes to neighborhoods
merged_neighborhoods <- neighborhoods_wgs84 %>%
  left_join(neighbourhood_stats, by = "Neighbourh")
merged_neighborhoods$ttc_rate = merged_neighborhoods$commute_ttc/merged_neighborhoods$population_2023
merged_neighborhoods$post_grad_prop = merged_neighborhoods$postsecondary_grad/merged_neighborhoods$population_2023
```

## Q1

```{r}
#6NN weight matrix
nc_centroids <- st_centroid(merged_neighborhoods)
coords <- st_coordinates(nc_centroids)
sids_kn6<-knn2nb(knearneigh(coords, k=6))
sids_kn6_w<-nb2listw(sids_kn6, style="W")

#Queen weight matrix
sids_nb_queen<-poly2nb(merged_neighborhoods, queen=TRUE)
sids_queen_w<-nb2listw(sids_nb_queen, style="W")

```

Queen lag

```{r}
#Queen lag
corrneigh_queen<-sp.correlogram(sids_nb_queen, merged_neighborhoods$median_family_income, order=10, method="I", style="W", randomisation=T, zero.policy=TRUE)
print(corrneigh_queen)
plot(corrneigh_queen,main="Moran's I for proportion of TTC communters, Correlogram, Queen Lags")
```

6NN lag

```{r}
corrneigh_6NN<-sp.correlogram(sids_kn6, merged_neighborhoods$median_family_income, order=10, method="I", style="W", randomisation=T, zero.policy=TRUE)
print(corrneigh_6NN)
plot(corrneigh_6NN,main="Moran's I for proportion of TTC communters, Correlogram, 6NN Lags")
```

We will choose Queen weight matrix, because it shows a larger Moran's I,
meaning it can capture the spatial autocorrelation better. Also, Queen
matrix highlights the spatial clustering of median family income, making
it a suitable choice for our spatial analysis.

## Q2 {.tabset}

### OLS

median_age

```{r}
mod_lm<-lm(median_family_income ~ post_grad_prop + unemploy_rate + Total_Popu + median_age + ttc_rate, data=merged_neighborhoods)
summary(mod_lm)
```

1.  **What:**This linear regression model does not account for spatial
    dependency. This model aims to describes the linear relationship
    between predictors and the response variable.

2.  **How:** The coefficients are estimated by minimizing the sum of the
    squared errors(the differences between the observed and predicted
    values). Also, the adjusted R-squared is 0.6582. Residual standard
    error is 20380, which measures the difference between observed and
    predicted values.

3.  **Description of parameters:**

    -   Intercept (1.783e+05): median family income is 1.783e+05 when
        all the predictors are zero

    -   unemploy_rate (-6.180e+03): For every 1-unit increase in the
        unemployment rate, median family income is estimated to decrease
        by 6.180e+03. It's statistically significant (p-value \< 0.05)

    -   Total_Popu (-5.439e-01): For every 1-unit increase in the
        unemployment rate, median family income is estimated to decrease
        by 5.439e-01. It's statistically significant (p-value \< 0.05)

    -   ttc_rate (-3.806e+05): For every 1-unit increase in the
        unemployment rate, median family income is estimated to decrease
        by 3.806e+05. It's statistically significant (p-value \< 0.05)

    -   post_grad_prop and median_age are not statistically significant
        predictors (p-value \> 0.05). Hence, we drop them out in the
        following models.

```{r}
merged_neighborhoods$lm_resid<-residuals(mod_lm)
# Check Moran's I for spatial autocorrelation in residuals (OLS model)
moran.test(merged_neighborhoods$lm_resid, sids_queen_w)
```

4.  **Test residual**

We can see Moran's I is positive(0.365375300), and statistically
significant(p_value \< 0.05). This suggests spatial autocorrelation is
present in the residuals, which means the OLS model does NOT fully
capture the spatial structure of the data. Also, the spatial
autocorrelation in the residuals can lead to biased estimates.

5.  **Map the residual**

```{r}
ggplot(merged_neighborhoods) + geom_sf(aes(fill = lm_resid))
```

### SAR Spatial Lag

```{r}
sar_lag_model <- lagsarlm(median_family_income ~  unemploy_rate + Total_Popu + ttc_rate, data=merged_neighborhoods, listw=sids_queen_w)
summary(sar_lag_model)
```

1.  **what:** The SAR Lag Model is fitted with the addition of a
    spatially lagged dependent variable to capture the spatial
    autocorrelation. SAR Lag Model only applies spatial weights to y. In
    other words, the errors is i.i.d

2.  **How**:

    1.  $Y = X\beta + \rho W Y + \epsilon$, W is the weights matrix,
        $\rho$ is the spatial autoregressive parameter, $\beta$ is the
        regression coefficients.

    2.  The goal is to maximize the likelihood:
        $L(\rho, \beta, \sigma^2 | Y, X) = -\frac{n}{2} \log(2 \pi \sigma^2) - \frac{1}{2 \sigma^2} \epsilon' \epsilon + \log |I - \rho W|$

    3.  $\beta$,$\rho$,$\sigma$ are optimized by Newton-Raphson

    4.  AIC is 3141.1, lower than OLS. A lower AIC indicates a better
        model fit.

3.  **Description of parameters:**

    -   Intercept (1.5783e+05): median family income is 1.5783e+05 when
        the predictors are zero

    -   Total_Popu (-4.1831e-01): It's NOT statistically significant
        (p-value \> 0.05), after accounting for spatial effects

    -   unemploy_rate (-4.7979e+03): It's statistically significant
        (p-value \< 0.05), after accounting for spatial effects

    -   ttc_rate (-4.0266e+05): It's statistically significant (p-value
        \< 0.05), after accounting for spatial effects

    -   The spatial parameter $\rho$ is 0.51351, and is statistically
        significant (p-value \< 0.005). This positive value indicates
        that neighborhoods with high/low median incomes are likely to be
        near other neighborhoods with high/low median incomes.

    -   The associated likelihood ratio test is 46.171 and statistically
        significant (p-value \< 0.05). This shows a higher likelihood
        for SAR Lag Model. This indicates the median income is fitted
        better by SAR Lag Model than the OLS model, because of the
        addition of the spatial autocorrelation.

4.  **Test residual**

```{r}
merged_neighborhoods$sar_lag_resid<-residuals(sar_lag_model)
moran.test(merged_neighborhoods$sar_lag_resid, sids_queen_w)
```

We can see Moran's I is positive(0.179115679), and statistically
significant(p_value \< 0.05). This suggests spatial autocorrelation is
present in the residuals, which means the SAR Spatial Lag model does NOT
fully capture the spatial structure of the data. However, the clustering
of similar residual by SAR Lag mode is much less than that by OLS model.
Hence, SAR Lag Model has effectively reduced spatial autocorrelation in
the residuals.

5.  **Map the residual**

```{r}
ggplot(merged_neighborhoods) + geom_sf(aes(fill = sar_lag_resid))
```

### SAR Spatial Error

```{r}
mod_sar_error <- spautolm(median_family_income ~ unemploy_rate + Total_Popu + ttc_rate, data=merged_neighborhoods, listw=sids_queen_w)
summary(mod_sar_error)
```

1.  **What:** $Y = X\beta + \epsilon$, where
    $\epsilon = \lambda W \epsilon + \nu$. $\lambda$ is the spatial
    autoregressive term for the errors, W is the spatial weight matrix,
    and ν is iid errors

2.  **How**:

    -   $Y \sim MVN(X\beta, \Sigma_{\epsilon})$, where
        $\Sigma_{\epsilon} = \sigma^2 (I - \lambda W)^{-1} (I - \lambda W)^{-T}$.

    -   The goal is to maximize the likelihood:
        $L(\lambda, \beta, \sigma^2 | Y, X) = -\frac{n}{2} \log(2 \pi \sigma^2) - \frac{1}{2 \sigma^2} (Y - X \beta)'(Y - X \beta) + \log |I - \lambda W|$

    -   $\beta$,$\rho$,$\sigma$ are optimized by Newton-Raphson

    -   AIC is 3135.9, lower than OLS and SAR Lag. A lower AIC indicates
        a better model fit.

3.  **Description of parameters**

    -   Intercept (2.2543e+05): median family income is 2.2543e+05, when
        the predictors are zero

    -   unemploy_rate (-5.2533e+03): It's statistically significant
        (p-value \< 0.05), after accounting for spatial effects

    -   Total_Popu (-2.8080e-01): It's NOT statistically significant
        (p-value \> 0.05), after accounting for spatial effects

    -   ttc_rate (-4.8373e+05): It's statistically significant (p-value
        \< 0.05), after accounting for spatial effects

    -   The spatial parameter $\lambda$ is 0.76599, and is statistically
        significant (p-value \< 0.005). This suggests there are
        unobserved spatial factors that are influencing median family
        income across neighborhoods.

    -   The associated likelihood ratio test is 51.401 and statistically
        significant (p-value \< 0.05). This shows a higher likelihood
        for SAR Error Model. This indicates the median income is fitted
        better by SAR Error Model than the OLS model, because of the
        addition of the error term that accounts for spatial
        autocorrelation.

4.  **Test residual**

```{r}
merged_neighborhoods$sar_error_resid<-residuals(mod_sar_error)
moran.test(merged_neighborhoods$sar_error_resid, listw=sids_queen_w)
```

We can see Moran's I is positive(0.065014886), and NOT statistically
significant(p_value \> 0.05). This suggests spatial autocorrelation is
NOT present in the residuals, which means the SAR Spatial Error model
does fully capture the spatial structure of the data. In other words, it
captures both the effects of the predictors and the spatially structured
unobserved factors that influence median family income across
neighborhoods.

5.  **Map the residual**

```{r}
ggplot(merged_neighborhoods) + geom_sf(aes(fill = sar_error_resid))
```

### SAR Spatial Lag-Error

```{r}
mod_sar_error_lag <- sacsarlm(median_family_income ~ unemploy_rate + Total_Popu + ttc_rate, data=merged_neighborhoods, listw=sids_queen_w)
summary(mod_sar_error_lag)
```

1.  **What:** $Y = X\beta + \epsilon$. The spatial autocorrelation in
    both $Y$,$\epsilon$ is accounted. we estimate both $\rho$ and
    $\lambda$.

2.  **How**:

    -   The goal is to maximize the likelihood:
        $L(\rho, \lambda, \beta, \sigma^2 | Y, X) = -\frac{n}{2} \log(2 \pi \sigma^2) - \frac{1}{2 \sigma^2} \epsilon' \epsilon + \log |I - \rho W| + \log |I - \lambda W|$

    -   $\beta$,$\rho$,$\sigma$ are optimized by Newton-Raphson

    -   AIC is 3127.8, lower than OLS, SAR Lag and SAR Error. A lower
        AIC indicates a better model fit.

3.  **Description of parameters**

    -   Intercept (1.8488e+05): median family income is 1.8488e+05, when
        all the predictors are zero

    -   unemploy_rate (-5.3155e+03): It's statistically significant
        (p-value \< 0.05), after accounting for spatial effects

    -   Total_Popu (-3.3095e-01): It's statistically significant
        (p-value \< 0.05), after accounting for spatial effects

    -   ttc_rate (-4.7106e+05): It's statistically significant (p-value
        \< 0.05), after accounting for spatial effects

    -   The spatial lag parameter $\rho$ is 0.36292, and is
        statistically significant (p-value \< 0.005). This positive
        value indicates that neighborhoods with high/low median incomes
        are likely to be near other neighborhoods with high/low median
        incomes.

    -   The spatial error parameter $\lambda$ is 0.53746, and is
        statistically significant (p-value \< 0.005). This suggests that
        there is spatial autocorrelation in the residuals. And there are
        unobserved spatial factors that are influencing median family
        income across neighborhoods.

    -   The associated likelihood ratio test is 61.493 and statistically
        significant (p-value \< 0.05). This shows a higher likelihood
        for SAR Lag-Error Model. This indicates the median income is
        fitted better by SAR Lag-Error Model than the OLS model, because
        it accounts for spatial autocorrelation both in the dependent
        variable and unobserved factors.

4.  **Test residual**

```{r}
merged_neighborhoods$sar_error_lag_resid<-residuals(mod_sar_error_lag)
moran.test(merged_neighborhoods$sar_error_lag_resid, listw=sids_queen_w)
```

We can see Moran's I is positive(0.028581170), and NOT statistically
significant(p_value \> 0.05). This suggests spatial autocorrelation is
NOT present in the residuals, which means the SAR Lag-Error model does
fully capture the spatial structure of the data. In other words, it
captures spatial autocorrelation in both the dependent variable and the
unobserved factors

5.  **Map the residual**

```{r}
ggplot(merged_neighborhoods) + geom_sf(aes(fill = sar_error_lag_resid))
```

### CAR

```{r}
#Queen weight matrix --- binary
sids_nb_queen<-poly2nb(merged_neighborhoods, queen=TRUE)
sids_queen_w_B<-nb2listw(sids_nb_queen, style="B")
```

```{r}
car_model <- spautolm(median_family_income ~ unemploy_rate + Total_Popu + ttc_rate, data=merged_neighborhoods, listw=sids_queen_w_B, family="CAR")
summary(car_model)
```

1.  **What:**
    $Y_i | Y_{-i} \sim N \left( \sum_{j \in \mathcal{N}_i} \beta_{ij} Y_j, \sigma_i^2 \right)$.
    it assumes that the state of a particular area is influenced only by
    its neighbors and not neighbors of neighbor. It assumes Markov
    random field property holds.

2.  **How**:

    -   Each observation Y(si) is conditional on the values of all other
        observations, $p[Y(s_i)\|Y_{−i}]$

    -   $\beta$,$\rho$,$\sigma$ are optimized by MLE and Newton-Raphson

    -   AIC is 3146.1, lower than OLS. A lower AIC indicates a better
        model fit.

3.  **Description of parameters**

    -   Intercept (2.2537e+05): median family income is 2.2537e+05, when
        all the predictors are zero

    -   unemploy_rate (-5.4707e+03): It's statistically significant
        (p-value \< 0.05), after accounting for spatial effects

    -   Total_Popu (-3.7846e-01): It's NOT statistically significant
        (p-value \> 0.05), after accounting for spatial effects

    -   ttc_rate (-4.7421e+05): It's statistically significant (p-value
        \< 0.05), after accounting for spatial effects

    -   The spatial error parameter $\lambda$ is 0.1468, and is
        statistically significant (p-value \< 0.005). This suggests that
        there is spatial autocorrelation in the residuals. And there are
        unobserved spatial factors that are influencing median family
        income across neighborhoods.

    -   The associated likelihood ratio test is 41.143 and statistically
        significant (p-value \< 0.05). This shows a higher likelihood
        for CAR Model. This indicates the median income is fitted better
        by CAR Model than the OLS model, because it accounts for spatial
        autocorrelation in the unobserved factors.

4.  **Test residual**

```{r}
merged_neighborhoods$car_resid<-residuals(car_model)

moran.test(merged_neighborhoods$car_resid, sids_queen_w)
```

We can see Moran's I is positive(0.024863018), and NOT statistically
significant(p_value \> 0.05). This suggests spatial autocorrelation is
NOT present in the residuals, which means the CAR model does fully
capture the spatial structure of the data. In other words, there is no
significant spatial autocorrelation in the residuals.

5.  **Map the residual**

```{r}
ggplot(merged_neighborhoods) + geom_sf(aes(fill = car_resid))
```

## Q3

We do not need to apply log transformation on the predictors

```{r}
hist(merged_neighborhoods$unemploy_rate)
```

```{r}
hist(merged_neighborhoods$Total_Popu)
```

```{r}
hist(merged_neighborhoods$ttc_rate)
```

### Queen neighborhood matrix with a CAR random effect

```{r}
merged_neighborhoods_drop <- merged_neighborhoods %>% st_drop_geometry()

spatial_corr <- as.matrix(nb2mat(sids_nb_queen, style = "B", zero.policy = TRUE))

rownames(spatial_corr) <- colnames(spatial_corr) <- merged_neighborhoods_drop$Neighbourh
```

```{r}
library(spaMM)
# Fit the model with a CAR random effect using the adjacency matrix
mixed_model_car <- fitme(
  median_family_income ~ unemploy_rate + Total_Popu + ttc_rate + adjacency(1 | Neighbourh),
  data = merged_neighborhoods_drop,
  family = gaussian(),
  adjMatrix = spatial_corr  
)
summary(mixed_model_car)
```

1.  **What:** The mixed effect model is $Y = X \beta + Z e + \epsilon$,
    where $e$\~$N(0, \Sigma_e)$ is the random effect, and Z accounts for
    the structure in the random effects.

2.  **How:**

    -   This model assume Each neighborhood(areal unit) has a random
        effect, and these effects are spatially correlated according to
        the Queen weight matrix
    -   The log-Likelihood value(-1567.073) is lower than SAR Lag-Error
        model (-1556.896). The higher the values (closer to zero) are,
        the better the model fit the data. So this is a worse fit than
        SAR Lag-Error.

3.  **Description of parameters:**

    -   Intercept (2.254e+05): median family income is 2.254e+05, when
        both post_grad_prop and unemploy_rate are zero

    -   post_grad_prop (-5.471e+03): It's statistically significant
        (t-value \< -1.96)

    -   unemploy_rate (-3.785e-01): It's statistically significant
        (t-value \< -1.96)

    -   unemploy_rate (-4.742e+05): It's statistically significant
        (t-value \< -1.96)

    -   The spatial parameter $\rho$ is 0.1467969 A positive value
        positive spatial autocorrelation, meaning that similar values in
        median family income are clustered togther.

    -   The residual variance(45882.8) is high. This suggests that there
        is still unexplained variability in the response after
        accounting for the fixed and random effects.

4.  **Test residuals:**

```{r}
merged_neighborhoods$mixed_model_car_resid<-residuals(mixed_model_car)
moran.test(merged_neighborhoods$mixed_model_car_resid, listw=sids_queen_w)
```

We can see Moran's I is negative(0.024897544), and NOT statistically
significant(p_value \> 0.05). This suggests spatial autocorrelation is
NOT present in the residuals, which means the this mixed effect model
does fully capture the spatial structure of the data.

5.  **Map the residuals:**

```{r}
ggplot(merged_neighborhoods) + geom_sf(aes(fill = mixed_model_car_resid))
```

### Matern adjacency with a neighborhood random effect

```{r}
# Fit the model with a Matern random effect based on geographical coordinates
centroids <- st_centroid(merged_neighborhoods)
coords <- st_coordinates(centroids)
merged_neighborhoods_drop$lon <- coords[, 1]
merged_neighborhoods_drop$lat <- coords[, 2]

mixed_model_matern <- fitme(
  median_family_income ~ unemploy_rate + Total_Popu + ttc_rate + Matern(1 | lon + lat),
  data = merged_neighborhoods_drop,
  family = gaussian()
)
```

```{r}
summary(mixed_model_matern)
```

1.  **What**: This mixed effect model have the fixed effects of
    unemploy_rate, Total_Popu, and ttc_rate. The random effect is a
    spatially structured effect using the Matern covariance function on
    longitude and latitude.

2.  **How**

    -   Each neighbour will have a random effect, and the Matern
        correlation will account for spatial distances for each
        neighbour.
    -   The log-Likelihood value(-1571.983) is lower than SAR Lag-Error
        model (-1556.896). The higher the values (closer to zero) are,
        the better the model fit the data. So this is a worse fit than
        SAR Lag-Error.

3.  **Description of data**

    -   Intercept (2.376e+05): median family income is 2.376e+05, when
        both post_grad_prop and unemploy_rate are zero

    -   post_grad_prop (-5.649e+03): It's statistically significant
        (t-value \< -1.96)

    -   unemploy_rate (-4.942e-01): It's statistically significant
        (t-value \< -1.96)

    -   unemploy_rate (-5.192e+05): It's statistically significant
        (t-value \< -1.96)

    -   the smoothness parameter, nu, is 0.6907037. This parameter
        controls the smoothness of the spatial correlation. A value
        close to 1 suggests a relatively smooth spatial variation in the
        random effects.

    -   The spatial parameter $\rho$ is 60.9986990. This implies that
        neighborhoods within this distance range have correlated values
        of median income. In other words, a larger $\rho$ indicates that
        observations far apart still exhibit some spatial correlation.

    -   The residual variance(30040400) is very high. This suggests that
        there is still unexplained variability in the response after
        accounting for the fixed and random effects.

4.  **Test residuals**

```{r}
merged_neighborhoods$mixed_model_matern_resid<-residuals(mixed_model_matern)
moran.test(merged_neighborhoods$mixed_model_matern_resid, listw=sids_queen_w)
```

We can see Moran's I is negative(0.010471305), and NOT statistically
significant(p_value \> 0.05). This suggests spatial autocorrelation is
NOT present in the residuals, which means the this mixed effect model
does fully capture the spatial structure of the data.

5.  **Map the residual**

```{r}
ggplot(merged_neighborhoods) + geom_sf(aes(fill = mixed_model_matern_resid))
```
